---
title: "*FinfoMDS*: Multidimensional scaling informed by *F*-statistic"
date: false
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
    number_sections: true
vignette: >
  %\VignetteIndexEntry{vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# About this vignette {.unnumbered}

This vignette describes how to compute the *F*-informed multidimensional scaling using the `FinfoMDS` package. `FinfoMDS` was developed by Soobin Kim ([sbbkim\@ucdavis.edu](mailto:sbbkim@ucdavis.edu){.email}). A proposal of the method and its full description can be found at:

-   H Kim⋆, S Kim⋆, JA Kimbrel, MM Morris, X Mayali and CR Buie (2025). Multidimensional scaling informed by *F*-statistic: Visualizing grouped microbiome data with inference, *arXiv* (<https://arxiv.org/abs/2308.00354v2>).

The vignette was last updated in May 2025.

# Introduction

Multidimensional scaling (MDS) is a dimensionality reduction technique used in microbial ecology data analysis to represent multivariate structures while preserving pairwise distances between samples. While its improvement has enhanced the ability to reveal data patterns by sample groups, these MDS-based methods often require prior assumptions for inference, limiting their broader application in general microbiome analysis.

Here, we introduce a new MDS-based ordination, ***F*****-informed MDS** (implemented in the R package `FinfoMDS`), which configures data distribution based on the *F*-statistic, the ratio of dispersion between groups that share common and different labels. Our approach offers a well-founded refinement of MDS that aligns with statistical test results, which can be beneficial for broader compositional data analyses in microbiology and ecology.

# Installation

## Bioconductor official release

To install an official release version of this package, start R (version "4.5") and enter:

```{r eval=FALSE}
BiocManager::install("FinfoMDS")
```

For older versions of R, please refer to the appropriate Bioconductor release.

## GitHub development version

The package may be updated before any changes migrate to the official release. The development version can be installed by entering:

```{r eval=FALSE}
devtools::install_github("soob-kim/FinfoMDS")
```

# Example

This section outlines steps for implementing `FinfoMDS` package to a microbiome dataset and how to obtain its 2D representation using *F*-informed MDS. Let's take an algal-associated bacterial community for example (Kim et al., 2022). First, load the example data by typing

```{r eval = FALSE}
data("microbiome", package = "FinfoMDS")
```

Next, compute the *F*-informed MDS by running:

```{r eval = FALSE}
result <- fmds(lambda = 0.3, D = microbiome$dist, y = microbiome$host)
```

.. where it will iterate until the 2D distributions converge or for 100 times by default—whichever occurs first. We have observed that setting `lambda` between 0.3 and 0.5 typically yields optimal results; however, the hyperparameter can be adjusted as long as it does not exceed 1.

The 2D representation of the community dataset is returned as a matrix and can be viewed by typing:

```{r eval = FALSE}
plot(result, pch=microbiome$host)
```

# Reference

H Kim, JA Kimbrel, CA Vaiana, JR Wollard, X Mayali, and CR Buie (2022). Bacterial response to spatial gradients of algal-derived nutrients in a porous microplate. *The ISME Journal*, **16**(4):1036–1045.

# Session information

```{r}
sessionInfo()
```
